
---
title: "EDA - Yellow 2023"
output: html_document
---

# General Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(vegan)
library(ggplot2)
```

# Load Data
```{r}
#data<- read.csv("/Users/cmantegna/Documents/GitHub/sicb_km/data/organisms_2023.csv")
data<- read.csv("/Users/cmantegna/Documents/GitHub/sicb_km/data/complete_data.csv")
```

# Check it out
```{r}
# does it look like it imported correctly? Check your row v column count in the environment window below: X observations (rows) of Y variables (columns).
head(data)

# take a look at your columns, their format, and look for any missing values (NAs), or overall inconsistencies in the data set before doing a deeper dive.
summary(data)

```

# Clean it up
## The current state of it isn't going to work, so we have to do some transformations:
1. Check for empty cells
2. Replace empties with the correct or imputed values
3. Consolidate rows so that there's a cumulative value for each zone in each section (only 3 rows per section)
3. Change the table format so that each section is formatted as a character
4. Change the table orientation so that the rows and columns are swapped; sections as columns and organisms as rows

```{r}

# check for empties
empty_cells <- sum(is.na(data) | data == "")
cat("Number of empty cells:", empty_cells, "\n")

#identify empties. this step is if you have really complicated data and have to manually check each one - we just need to replace empties with 0's.
rows_with_empty <- which(rowSums(is.na(data) | data == "") > 0)
cols_with_empty <- which(colSums(is.na(data) | data == "") > 0)

# replace empties with 0 & then rerun the empty cell step above to double check they were all replaced
data[is.na(data) | data == ""] <- 0

#overwrite your data file so we don't have to do this again and write it out as a safety stop.
data <- data
write_csv(data, "/Users/cmantegna/Documents/GitHub/sicb_km/data/complete_data.csv")

```

# Consolidate counts

```{r}

# make sections characters & check they really are characters
data$SECTION <- as.character(data$SECTION)
str(data)

# we're using dplyr
library(dplyr)

# Summarize data to a single cumulative value for each section and zone, check it worked and write it out for your second safety stop.
data <- data %>%
  group_by(SECTION, ZONE) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

head(data)

write_csv(data, "/Users/cmantegna/Documents/GitHub/sicb_km/data/cleaned_data.csv")

```

# Transform data set
```{r}

# Load necessary libraries
library(readxl)
library(tidyr)
library(dplyr)

# Pivot 
pivoted_data <- data %>%
  pivot_longer(
    cols = -c(SECTION, ZONE),       # Columns to keep fixed
    names_to = "Organism",          # Name for the new variable column
    values_to = "Count"             # Name for the new value column
  ) %>%
  pivot_wider(
    names_from = SECTION,           # Make sections into columns
    values_from = Count             # Values to fill in the new columns
  )

# check your work
head(pivoted_data)

# write out your data as another safety stop
write.csv(pivoted_data, "/Users/cmantegna/Documents/GitHub/sicb_km/data/pivoted_data.csv", row.names = FALSE)

```


