---
title: "Analysis - Yellow 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r}

library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load Data
```{r}

# figure out where you are on your computer
getwd()

# Note you'll need to change the file path to match where it is on your computer
data<- read.csv("/Users/cmantegna/Documents/GitHub/sicb_km/output/data_adjusted/cleaned_data.csv")
pivoted_data<- read.csv("/Users/cmantegna/Documents/GitHub/sicb_km/output/data_adjusted/pivoted_data.csv")
```

# Community composition analysis
## shanon diversity index
### these results follow what we would expect.
```{r}

# Shannon Diversity Index. This measures the richness (# of different species present) and evenness of species distribution. Higher numbers indicate greater diversity and even distribution, lower numbers indicate lower diversity d/y a few species dominating the landscape.
library(tibble)

# SDI calculation
diversity_indices <- apply(select(pivoted_data, -organism), 2, function(x) diversity(x, index = "shannon"))

# convert to df for ease of viewing
diversity_table <- tibble(
  site = names(diversity_indices),
  Shannon_Index = diversity_indices
)

# check it out + write it out
print(diversity_table)

write.csv(diversity_table, "/Users/cmantegna/Documents/GitHub/sicb_km/output/tables/sdi.csv", row.names = FALSE)

```

## SDI heatmap - not helpful at all
```{r}

library(ggplot2)

# Add zones for labeling
diversity_table <- diversity_table %>%
  mutate(zone = substr(site, nchar(site), nchar(site)))

# Plot heatmap
ggplot(diversity_table, aes(x = zone, y = site, fill = Shannon_Index)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Shannon Diversity Index Heatmap",
    x = "Zone",
    y = "Site",
    fill = "Diversity Index"
  ) +
  theme_minimal()

```

## SDI barplot - this confirms what should have happened; viz not useful
```{r}

ggplot(diversity_table, aes(x = zone, y = Shannon_Index, fill = zone)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Shannon Diversity Index by Zone",
    x = "Zone",
    y = "Diversity Index"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")

```

## general heatmap - terrible
```{r}

#install.packages('pheatmap')
library(pheatmap)

# Create heatmap
pheatmap(as.matrix(select(pivoted_data, -organism)),
         cluster_rows = TRUE, cluster_cols = TRUE,
         main = "Heatmap of Organism Counts Across Zones",
         color = colorRampPalette(c("white", "blue"))(50))

```

## simpson index for dominance (by section and zone)
### Interpretation: Simpson values are between 0-1; 1 means only one species is dominanting the environment and 0 means there's unlimited diversity. Lower numbers indicate higher diversity.
### Interpretation: Species richeness is just the unique number of species identified in each.
```{r}

# create community matrix
community_matrix <- pivoted_data %>%
  column_to_rownames("organism") %>% # Set Organism as row names
  t() %>% # Transpose to make species columns
  as.data.frame()

# fix df so the sites are part of the df
community_matrix <- community_matrix %>%
  rownames_to_column(var = "site") 

#write.csv(community_matrix, "/Users/cmantegna/Documents/GitHub/sicb_km/output/tables/community_matrix.csv", row.names = FALSE)

# Simpson Index
simpson_index <- function(row) {
  total <- sum(row)
  if (total == 0) return(0) # Avoid division by zero
  proportions <- row / total
  return(sum(proportions^2))
}

# Species Richness
species_richness <- function(row) {
  return(sum(row > 0))
}

# create summary table of simpson and richness
summary_table <- community_matrix %>%
  rowwise() %>%
  mutate(
    Simpson_Index = simpson_index(c_across(-site)),
    Species_Richness = species_richness(c_across(-site))
  ) %>%
  select(site, Simpson_Index, Species_Richness) 

# check it and write out to file
print(summary_table)

write.csv(summary_table, "/Users/cmantegna/Documents/GitHub/sicb_km/output/tables/simpson_summary.csv", row.names = FALSE)

```

## simpson index for dominance (separately by section and zone)
### create an adjusted df
```{r}

library(stringr)

# add columns for section and zone using regex
community_matrix_sep <- community_matrix %>%
  mutate(
    Section = str_extract(site, "\\d+"),  # Extract the numeric part
    Zone = str_extract(site, "[A-Z]$")   # Extract the last uppercase letter
  )

```

## simpson index for dominance by zone
```{r}

library(dplyr)
library(stringr)


# simpson and richness
community_matrix_sep <- community_matrix_sep %>%
  rowwise() %>%
  mutate(
    Simpson_Index = sum((c_across(-c(site, Section, Zone)) / sum(c_across(-c(site, Section, Zone))))^2, na.rm = TRUE),
    Species_Richness = sum(c_across(-c(site, Section, Zone)) > 0, na.rm = TRUE)
  ) %>%
  ungroup()

# calculate by zone
summary_by_zone <- community_matrix_sep %>%
  group_by(Zone) %>%
  summarise(
    Mean_Simpson = mean(Simpson_Index, na.rm = TRUE),
    Mean_Richness = mean(Species_Richness, na.rm = TRUE)
  )

# View the results
print(summary_by_zone)


```

## simpson index for dominance by section
```{r}

# Group by Section and calculate metrics
summary_by_section <- community_matrix_sep %>%
  group_by(Section) %>%
  summarise(
    Mean_Simpson = mean(Simpson_Index, na.rm = TRUE),
    Mean_Richness = mean(Species_Richness, na.rm = TRUE)
  )

# View the results
print(summary_by_section)

```

## nmds
### Interpretation - zonation cluster and section clustering tells who is most similar in composition. None of the clustering follows what we would expect from a substrate pov.
```{r}

library(vegan)

# Calculate Bray-Curtis Dissimilarity
bray_curtis <- vegdist(community_matrix, method = "bray")

# Perform NMDS
nmds <- metaMDS(bray_curtis)

# Plot NMDS Results
plot(nmds, type = "t", main = "NMDS of Community Composition")

```

# #stress plot - i will need to dig in further to understand why we have a distinct stair-step pattern
```{r}

stressplot(nmds)

```

## simpson and richness based on groups
### Interpretation: neither substrate nor water movement solely impact community diversity or species richness.
```{r}

# We want to look at the difference substrate and water movement direction play on diversity and richness, so we'll create 3 groups (section 1-3, 4-5, and 6-8). Since sections 1-3 and 6-8 share the same substrate, but differ in ordinal orientation, we'll separate them into two groups.

# create custom groups
community_matrix_groups <- community_matrix_sep %>%
  mutate(
    Section_Group = case_when(
      as.numeric(Section) %in% 1:3 ~ "1-3",
      as.numeric(Section) %in% 4:5 ~ "4-5",
      as.numeric(Section) %in% 6:8 ~ "6-8",
      TRUE ~ "Other" # Catch any unexpected values
    )
  )

# Group by the new Section_Group and calculate metrics
summary_by_section_group <- community_matrix_groups %>%
  group_by(Section_Group) %>%
  summarise(
    Mean_Simpson = mean(Simpson_Index, na.rm = TRUE),
    Mean_Richness = mean(Species_Richness, na.rm = TRUE)
  )

# View the results
print(summary_by_section_group)

```

## permanova setup
```{r}

# set it up
community_data <- community_matrix_groups %>% select(-site, -Zone, -Section, -Simpson_Index, -Species_Richness, -Section_Group) # only species columns
metadata <- community_matrix_groups %>% select(site, Zone, Section, Section_Group) # Grouping variables

# calculate dissimilarity
bray_curtis <- vegdist(community_data, method = "bray")

```

## permanova - validate
### Interpretation: Zone explains a statistically significant amount of variation in the community composition where neither section nor section group do.
```{r}

# zone
# result: p= 0.001 ***
# result: explains about 32% of the differences
permanova_zone <- adonis2(bray_curtis ~ Zone, data = metadata, permutations = 999)
print(permanova_zone)

# section
# result: no significant differences
permanova_section <- adonis2(bray_curtis ~ Section, data = metadata, permutations = 999)
print(permanova_section)

# section
# result: no significant differences
permanova_section_group <- adonis2(bray_curtis ~ Section_Group, data = metadata, permutations = 999)
print(permanova_section_group)

```

## permanova with interactions - validate
### Interactions
```{r}

# zone + section
# result: no significant difference
permanova_zs <- adonis2(bray_curtis ~ Zone * Section, data = metadata, permutations = 999)
print(permanova_zs)

# zone + section group
# result: p= 0.001 ***
# result: explains about 53% of the differences
permanova_zsg <- adonis2(bray_curtis ~ Zone * Section_Group, data = metadata, permutations = 999)
print(permanova_zsg)

```

plot nmds of interaction
```{r}

plot(nmds, type = "t")
ordihull(nmds, groups = interaction(metadata$Zone, metadata$Section_Group), draw = "polygon", col = "blue", label = TRUE)

```

## new glm
```{r}

library(mgcv)

# Generalized Additive Model (GAM)
gam_model <- gam(Count ~ s(Temperature) + s(Depth), data = long_data, family = poisson)
summary(gam_model)

```






